Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-01-23 src/tex/ms.tex
[showyourwork] overleaf sync

\textcolor{red}{see philcox results for boiler plate text...}

We test our analysis using mock observations drawn firstly from \texttt{CHEMPY} from the same yield set as the training data, secondly using a different yield set to test for potential biases through model missspecification in our SBI setup and then lastly from a large-scale hydrodynamical simulation taken from the IlustrisTNG suite \citep{Pillepich2018} but with the same yield set as our \texttt{CHEMPY} training data to ensure that we recover the correct parameters even for models with a completely different treatment of ISM physics. The methods presented here can naturally be extended to any fast and flexible GCE model, not just \texttt{Chempy}.

\subsection{\texttt{CHEMPY} Mock Observational Data}
Our analysis uses mock observations drawn from the \textit{neural network} emulator with fixed values of the global galactic parameters $\alpha_{\rm IMF}=-2.3$ and $\log_{10}(N_{\rm Ia})=-2.89$ and the local parameters $\vec{\Theta}_i$ from Tab. \ref{tab:priors}, additionally drawing $T_i$ uniformly in the range $[2,12.8]$ Gyr, to minimize overlap with the neural network training birth-time limits when observational uncertainties are included. 
Each set of parameters is passed to the NN, producing eight true chemical abundances, to which we add a Gaussian error of $0.05$ dex.

@@

\paragraph{Validation through absolute percentage error}
To evaluate our method quantitatively, we compare the posterior mean to the ground truth value for each observation and calculate the absolute percentage error (APE, see Fig.~\ref{fig:posterior_APE}). For a single star observation, our NPE has an APE of \variable{output/global_posterior_APE.txt} when looking at all 6 parameters of interest. If we restrict ourselves to only the global two parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$ our NPE achieves an APE of \variable{output/posterior_APE.txt} as we show in Fig.~\ref{fig:posterior_APE}. Looking at Fig.~\ref{fig:posterior_APE} we find that for an individual star the accuracy of the NPE is not particularly high. However, for the global parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$ we can boost the accuracy by combining the inference for many stars of the same galaxy.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/CHEMPY_TNG_yields_Nstar_comp.pdf}
    \vspace{-.25cm}
    \caption{Accuracy of inferred global galactic parameters $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ as a function of number of observed stars, comparing SBI (blue) to the inferred values using HMC (red) as done by \cite{Philcox_2019}.}
    \label{fig:CHEMPY_TNG_N_star_analysis}
    \script{chempy_tng_inference.py}
\end{figure}

@@

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/CHEMPY_TNG_yields.pdf}
    \vspace{-.25cm}
    \caption{Sampled posteriors for global galactic parameters $\alpha_{\rm IMF}$ and $\log_{10}(\rm N_{Ia})$ of $1000$ stars.}
    \label{fig:CHEMPY_TNG_sbi} 
    \script{chempy_tng_inference.py}
\end{figure}

%
Taking the means of multiple observed stars gives us higher accuracy and precision of the global galactic parameters $\vec\Lambda$ (Fig. \ref{fig:N_star_analysis}). However, we also see that in the limit of very few stars (less than $\sim10$) SBI shows a larger bias than HMC results. But in the limit of large numbers of stars (few hundred) the accuracy and precision is superior compared to HMC.
%
To infer the global galactic parameters $\vec\Lambda$ of the stars as accurate as possible we take the mean of all inferred posteriors from $n_{\rm stars}=1000$ (Fig. \ref{fig:sbi}) resulting in $\alpha_{\rm IMF}=-2.292\pm0.003$ and $\log_{10}(\rm N_{Ia})=-2.883\pm0.005$ which deviates less than $0.4\%$ from the correct value.

>>>>

We use our SBI method described in the previous section to infer the global galactic parameters $\alpha_{\rm {IMF}}$ and $\log_{10}(N_{\rm{Ia}})$. In order to demonstrate the performance and robustness of our methods we use three mock data-sets: 
\begin{enumerate}
    \item Mock observations drawn from \texttt{CHEMPY} from the same yield set as the training data for the neural network emulator. With this we esnure to test our training strategy and the performance without any systematic distribution shifts. 
    \item  \texttt{CHEMPY} mock data using a different yield set to test for potential biases through model missspecification in our SBI setup.
    \item Simulated data from a large-scale hydrodynamical simulation taken from the IlustrisTNG suite \citep{Pillepich2018} but with the same yield set as our \texttt{CHEMPY} training data to ensure that we recover the correct parameters even for models with a completely different treatment of ISM physics.
\end{enumerate}

Note, the methods presented here can naturally be extended to any fast and flexible GCE model, not just \texttt{Chempy}.

\subsection{\texttt{CHEMPY} Mock Observational Data}
Our analysis uses mock observations drawn from the \textit{neural network} emulator with fixed values of the global galactic parameters $\alpha_{\rm IMF}=-2.3$ and $\log_{10}(N_{\rm Ia})=-2.89$ and varying local parameters $\vec{\Theta}_i$ using priors from Tab. \ref{tab:priors}. Additionally we draw $T_i$ uniformly in the range $[2,12.8]$ Gyr to minimize overlap with the neural network training birth-time limits when observational uncertainties are included. 
Each set of parameters is passed to our \texttt{CHEMPY} emulator, producing eight true chemical abundances. In order to fully represent observational data, we augment this data with observational uncertainties by adding a Gaussian error of $0.05$ dex for the abundances representative for typical APOGEE data \citep{Majewski2016}. All our datasets and accompanying code to reproduce our data generation process and analysis is publicly available\footnote{\url{https://github.com/TobiBu/sbi-chempy}}.

@@

\paragraph{Validation through absolute percentage error}
We start evaluating our method by using \texttt{CHEMPY} to produce a mock observational dataset to ensure no systematical shft between training and testing data in terms of physics parameters.

To evaluate our method quantitatively, we compare the posterior mean to the ground truth value for each observation and calculate the absolute percentage error (APE, see Fig.~\ref{fig:posterior_APE}). For a single star observation, our NPE has an APE of \variable{output/global_posterior_APE.txt} when looking at all 6 parameters of interest. If we restrict ourselves to only the global two parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$ our NPE achieves an APE of \variable{output/posterior_APE.txt} as we show in Fig.~\ref{fig:posterior_APE}. Looking at Fig.~\ref{fig:posterior_APE} we find that for an individual star the accuracy of the NPE is not particularly high. However, for the global parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$ we can boost the accuracy by combining the inference for many stars of the same galaxy.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/CHEMPY_TNG_yields_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Accuracy of inferred global galactic parameters $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ as a function of number of observed stars, comparing SBI (blue line) to the inferred values using HMC (red line) as done by \cite{Philcox_2019} and the ground truth values (black dashed line). For the SBI analysis we show $1\sigma$ and $2\sigma$ contours while HMC results only show $1\sigma$ statistical uncertainities as reported in Tab.~3 of \citet{Philcox_2019} (blue/red shaded regions).}
    \label{fig:CHEMPY_TNG_N_star_analysis}
    \script{chempy_tng_inference.py}
\end{figure*}

@@

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/CHEMPY_TNG_yields.png}
    \vspace{-.25cm}
    \caption{Joint posterior for global galactic parameters $\alpha_{\rm IMF}$ and $\log_{10}(\rm N_{Ia})$ of $1000$ stars. The ground truth value is shown by the black dot, the posterior mean of the SBI inference is shown with the red star and white and black ellipses show the $1-3\sigma$ contours of our inference.}
    \label{fig:CHEMPY_TNG_sbi} 
    \script{chempy_tng_inference.py}
\end{figure}

%
Combining the inference on multiple observed stars gives us higher accuracy and precision of the global galactic parameters $\vec\Lambda$. We therefore perform inference using a range of stars $n\in[1,1000]$ (see Fig. \ref{fig:CHEMPY_TNG_N_star_analysis}) to show how inference accuracy increases with number of observations. We find that in the limit of few stars (less than $\sim100$) SBI shows a larger uncertainties than the HMC results of \citet{Philcox_2019}. But in the limit of large numbers of stars (few hundred) the accuracy and precision is superior compared to HMC. In particular, after using a few ten stars, the NPE estimate is less biased than the HMC results. 
Finally, given our computational advantage we will be able to use orders of magnitude more stars for our inference.
This is particular important since sample variances play an important role when using small samples of stars as also noted in \citet{Philcox_2019}.
%

In Fig.~\ref{fig:CHEMPY_TNG_sbi} we show the joint posterior for $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ for our inference. The red star indicates the ground truth value, the black dot shows our posterior mean and white contours indicate $1-3\sigma levels$.

Using a sample of 1000 stars we infer \variable{output/CHEMPY_TNG_sbi.txt} which deviates less than $\sim0.04\%$ from the correct value.

Note, as shown in Fig.~\ref{fig:posterior_APE} in principle our analysis is also able to infer the local parameters $\Vec{\Theta}_i$ and $T_i$. This would allow us to estimate/infer stellar ages as well. However, currently our NPE network is not particularly good at estimating ages from abundances alone. On avergae our inference for ages results in an APE of $\sim27\%$ which is slightly larger than the observational noise of $20\%$ that we add during mock up of our data.

In summary, our SBI pipeline is quite capable of correctly and precisely inferring  global parameters of chemical enrichment models from stellar abundance alone when using the same physical model and yield tables as during training. Next, we will check what happens if for some reason the training data was generated with a different yield set.
